\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-09-18 paper/ms.tex
different page breaks for pdflatex
removing response text
some responding to referee

\newcommand{\response}[1]{{\color{blue}#1}}

@@

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
\response{The semiseparable nature of these matrices was first recognized by
\citet{Ambikasaran:2015}, building on intuition from a twenty year old paper
\citep{Rybicki:1995}.}
As we will discuss in the following pages, this choice of kernel function
arises naturally in physical systems and we demonstrate it can be used as
an effective model in other cases.
This method is especially appealing compared to other similar methods~--~we
discuss these comparisons in \sect{compare}~--~because it is exact, simple,
and fast.

@@

GPs are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ parameterized by
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, \response{the log-likelihood function $\mathcal{L}
(\bvec{\theta},\,\bvec{\alpha})$} with a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln \mathcal{L} (\bvec{\theta},\,\bvec{\alpha}) =
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
\eq{gp-likelihood} is the equation of an $N$-dimensional Gaussian and it can
be derived as the generalization of what we astronomers call ``$\chi^2$''
to include the effects of correlated noise.
\response{
A point estimate for the values of the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.}
\response{Furthermore, the uncertainties on $\bvec{\theta}$ and
$\bvec{\alpha}$ can be quantified by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.}

@@

Similarly, setting both $b_j$ and $d_j$ to zero, we get
\begin{eqnarray}
k_j(\tau_{nm}) = a_j\,\exp\left(-c_j\,\tau_{nm}\right)
\end{eqnarray}
with the PSD
\begin{eqnarray}
S_j(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{a_j}{c_j}\,
    \frac{1}{1+{\left(\frac{\omega}{c_j}\right)}^2} \quad.
\end{eqnarray}
This model is often called an Ornstein--Uhlenbeck process \response{\citep[in
reference to the classic paper,][]{Uhlenbeck:1930}} and this is the kernel
that was studied by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.

Finally, we note that the product of two terms of the form found inside the
sum in \eq{celerite-kernel} can also be re-written as a sum with updated
parameters
\begin{eqnarray}\eqlabel{product-rule}
k_j(\tau) \, k_k(\tau) =
    e^{-\tilde{c}\,\tau}\,[
        \tilde{a}_+\,\cos(\tilde{d}_+\,\tau) + \tilde{b}_+\,\sin(\tilde{d}_+\,\tau) +
        \tilde{a}_-\,\cos(\tilde{d}_-\,\tau) + \tilde{b}_-\,\sin(\tilde{d}_-\,\tau)
    ]
\end{eqnarray}
where
\begin{eqnarray}
    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
    \tilde{c} &=& c_j + c_k \\
    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
\end{eqnarray}
Therefore, the method described in \sect{factor} can be used to
perform scalable inference on large datasets for any model, where the kernel
function is a sum or product of \celeriteterm\ terms.

@@

The relationship between the \celeriteterm\ model and semiseparable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015} building
on earlier work by Rybicki \& Press \citep{Rybicki:1995}.
Ambikasaran derived a scalable direct solver for any general semiseparable
matrix and applied this method to \celeriteterm\ models with $b_j = 0$ and
$d_j = 0$.
During the preparation of this paper, we generalized this earlier method to
include non-zero $b_j$ and $d_j$, but it turns out that we can derive a higher
performance algorithm by restricting our method to positive definite matrices.
In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
positive definite.

@@

\subsection{How to choose a celerite model}

@@

For a similar class of models~--~we return to these below in
\sect{compare}~--~\citet{Kelly:2014} recommend using an ``information
criterion'', such as the Akaike information criterion (AIC) or the Bayesian
information criterion (BIC), as a model selection metric.
These criteria are easy to implement and we use the BIC in \sect{example2}
but, since these are only formally valid under specific sets of assumptions, we
cannot recommend their use in general.
}

\newpage
\section{Examples}\sectlabel{examples}

@@

The next three examples show \celerite\ used to infer the properties of stars
and exoplanets observed by NASA's \kepler\ Mission.
Each of these examples touches on an active area of research so we limit our
examples to be qualitative in nature and do not claim that \celerite\ is the
optimal method, but we hope these examples encourage interested readers to
investigate the applicability of \celerite\ to their research.

In each example, the joint posterior probability density is given by
\begin{eqnarray}\eqlabel{joint-post}
p(\bvec{\theta},\,\bvec{\alpha}\,|\,\bvec{y},\,X) \propto
p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})\,
p(\bvec{\theta},\,\bvec{\alpha})
\end{eqnarray}
where $p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})$ is the GP
likelihood defined in \eq{gp-likelihood} and $p(\bvec{\theta},\,\bvec{\alpha})$
is the joint prior probability density for the parameters.
We sample each posterior density using MCMC and investigate the performance of
the model when used to interpret the physical processes that generated the
data.
The specific parameters and priors are discussed in detail in each section,
but we generally assume separable uniform priors with plausibly broad support.
}

@@

To start, we estimate the maximum \emph{a posteriori} (MAP) parameters using
the \project{L-BFGS-B} non-linear optimization routine \citep{Byrd:1995,
Zhu:1997} implemented by the \project{SciPy} project \citep{Jones:2001}.
The top left panel of \Figure{simulated-correct} shows the conditional mean
and standard deviation of the MAP model over-plotted on the simulated data and
the bottom panel shows the residuals away from this model.
Since we are using the correct model to fit the data, it is reassuring that
the residuals appear qualitatively uncorrelated in this figure.

We then sample the joint posterior probability (\eqalt{joint-post}) using
\emcee\ \citep{Goodman:2010, Foreman-Mackey:2013}.
We initialize \exampleinwalkers~walkers by sampling from a three-dimensional
isotropic Gaussian centered the MAP parameter vector and with a standard
deviation of $10^{-4}$ in each dimension.
We then run \exampleinburn~steps of burn-in and \exampleinsteps~steps of MCMC.
To assess convergence, we estimate the mean integrated autocorrelation time of
the chain across parameters \citep{Sokal:1989, Goodman:2010} and find that the
chain results in \exampleineff~effective samples.

@@

\newpage
\subsection{Example 4: Asteroseismic oscillations}\sectlabel{astero}

@@

\subsection{Code availability}

@@

For a GP kernel to be valid, it must produce a positive definite covariance
matrix for all input coordinates.
For stationary kernels, this is equivalent~--~by Bochner's theorem \citep[see
Section 4.2.1 in][]{Rasmussen:2006}~--~to requiring that the kernel be the
Fourier transform of a positive finite measure.
This means that the power spectrum of a positive definite kernel must be
positive for all frequencies.
This result is intuitive because, since the power spectrum of a process is
defined as the expected squared amplitude of the Fourier transform of the time
series, it must be non-negative.

>>>>

\newcommand{\response}[1]{{#1}}
\newcommand{\responseb}[1]{{\color{blue}{#1}}}

@@

This method achieves linear scaling by exploiting structure in the covariance
matrix when it is generated by a mixture of exponentials.
\response{The semiseparable nature of these matrices was first recognized by
\citet{Ambikasaran:2015}, building on intuition from a twenty year old paper
\citep{Rybicki:1995}.}
As we will discuss in the following pages, this choice of kernel function
arises naturally in physical systems and we demonstrate it can be used as
an effective model\footnote{\responseb{Here and throughout, we use the term
``effective model'' to refer to a model that we use to make inferences even
though it is known to be wrong.}} in other cases.
This method is especially appealing compared to other similar methods~--~we
discuss these comparisons in \sect{compare}~--~because it is exact, simple,
and fast.

@@

GPs are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ parameterized by
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, \response{the log-likelihood function $\mathcal{L}
(\bvec{\theta},\,\bvec{\alpha})$} with a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln \mathcal{L} (\bvec{\theta},\,\bvec{\alpha}) =
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
\eq{gp-likelihood} is the equation of an $N$-dimensional Gaussian and it can
be derived as the generalization of what we astronomers call ``$\chi^2$''
to include the effects of correlated noise.
\response{
A point estimate for the values of the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood}~--~\responseb{or the likelihood times a prior for
a Bayesian estimate}~---~with respect to $\bvec{\theta}$ and $\bvec{\alpha}$
using a non-linear optimization routine \citep{Nocedal:2006}.}
\responseb{Furthermore, the posterior probability density for $\bvec{\theta}$
and $\bvec{\alpha}$ can be quantified by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.}

@@

Similarly, setting both $b_j$ and $d_j$ to zero, we get
\begin{eqnarray}
k_j(\tau_{nm}) = a_j\,\exp\left(-c_j\,\tau_{nm}\right)
\end{eqnarray}
with the PSD
\begin{eqnarray}
S_j(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{a_j}{c_j}\,
    \frac{1}{1+{\left(\frac{\omega}{c_j}\right)}^2} \quad.
\end{eqnarray}
This model is often called a \responseb{``damped random walk'' or}
Ornstein--Uhlenbeck process \response{\citep[in reference to the classic
paper,][]{Uhlenbeck:1930}} and this is the kernel that was studied by Rybicki
\& Press \citep{Rybicki:1992,Rybicki:1995}.

Finally, we note that the product of two terms of the form found inside the
sum in \eq{celerite-kernel} can also be re-written as a sum with updated
parameters
\begin{eqnarray}\eqlabel{product-rule}
k_j(\tau) \, k_i(\tau) =
    e^{-\tilde{c}\,\tau}\,[
        \tilde{a}_+\,\cos(\tilde{d}_+\,\tau) + \tilde{b}_+\,\sin(\tilde{d}_+\,\tau) +
        \tilde{a}_-\,\cos(\tilde{d}_-\,\tau) + \tilde{b}_-\,\sin(\tilde{d}_-\,\tau)
    ]
\end{eqnarray}
where
\begin{eqnarray}
    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_i \pm b_j\,b_i) \\
    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_i \mp a_j\,b_i) \\
    \tilde{c} &=& c_j + c_i \\
    \tilde{d}_{\pm} &=& d_j \mp d_i \quad.
\end{eqnarray}
Therefore, the method described in \sect{factor} can be used to
perform scalable inference on large datasets for any model, where the kernel
function is a sum or product of \celeriteterm\ terms.

@@

The relationship between the \celeriteterm\ model and semiseparable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015} building
on earlier work by Rybicki \& Press \citep{Rybicki:1995}.
Ambikasaran derived a scalable direct solver for any general semiseparable
matrix and applied this method to \celeriteterm\ models with $b_j = 0$ and
$d_j = 0$.
\responseb{The \citet{Ambikasaran:2015} method uses a factorization routine
that can be applied to general semiseparable matrices.
In other words, it can be applied to factorize matrices even if they are not
positive definite.
While preparing this paper, we generalized this method to include non-zero
$b_j$ and $d_j$, but all valid GP covariance matrices are symmetric and
positive definite so we can exploit this constraint to derive a Cholesky
factorization algorithm.
It turns out that this algorithm is more than an order of magnitude faster
than the previously published method when applied to \celeriteterm\ models.
In this and the following sections, we derive this novel factorization
algorithm and, in \app{psd}, we discuss methods for ensuring that the
covariance matrix for a \celeriteterm\ model is positive definite.}

@@

\newpage
\subsection{How to choose a celerite model}

@@

For a similar class of models~--~we return to these below in
\sect{compare}~--~\citet{Kelly:2014} recommend using an ``information
criterion'', such as the Akaike information criterion (AIC) or the Bayesian
information criterion (BIC), as a model selection metric.}
\responseb{These criteria are easy to implement and we use the BIC in
\sect{example2}, but each information criterion comes with caveats that should
be considered in any application \citep[see, for example,][]{Ivezic:2014}.}

\section{Examples}\sectlabel{examples}

@@

The next three examples show \celerite\ used to infer the properties of stars
and exoplanets observed by NASA's \kepler\ Mission.
Each of these examples touches on an active area of research so we limit our
examples to be qualitative in nature and do not claim that \celerite\ is the
optimal method, but we hope these examples encourage interested readers to
investigate the applicability of \celerite\ to their research.}

\response{
In each example, the joint posterior probability density is given by
\begin{eqnarray}\eqlabel{joint-post}
p(\bvec{\theta},\,\bvec{\alpha}\,|\,\bvec{y},\,X) \propto
p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})\,
p(\bvec{\theta},\,\bvec{\alpha})
\end{eqnarray}
where $p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})$ is the GP
likelihood defined in \eq{gp-likelihood} and $p(\bvec{\theta},\,\bvec{\alpha})$
is the joint prior probability density for the parameters.
We sample each posterior density using MCMC and investigate the performance of
the model when used to interpret the physical processes that generated the
data.
The specific parameters and priors are discussed in detail in each section,
but we generally assume \responseb{independent} uniform priors with plausibly
broad support.}

\responseb{
While we do perform a convergence analysis in each case by computing the
autocorrelation function and integrated autocorrelation time for the target
chain \citep{Sokal:1989, Goodman:2010}, the usual caveats associated with MCMC
analysis apply when interpreting these results.
For example, if there are multiple posterior modes that are not sampled by any
chain, the autocorrelation analysis will not identify that the sampler has not
converged.
There are published sampling algorithms \citep{Gregory:2005, Liu:2008,
Feroz:2009, Brewer:2011, Kelly:2014} that can, in some cases, be less
sensitive to multi-modality than the sampler that we use
\citep{Goodman:2010,Foreman-Mackey:2013}, but a comparison of sampling
algorithms is beyond the scope of this paper and \celerite\ will also improve
the computational efficiency of GP models when using the other inference
algorithms.
}

@@

To start, we estimate the maximum \emph{a posteriori} (MAP) parameters using
the \project{L-BFGS-B} non-linear optimization routine \citep{Byrd:1995,
Zhu:1997} implemented by the \project{SciPy} project \citep{Jones:2001}.
The top left panel of \Figure{simulated-correct} shows the mean and standard
deviation of the \responseb{predicted noise-free data distribution from the
MAP model over-plotted as a blue contour} on the simulated data and the bottom
panel shows the residuals away from this model.
Since we are using the correct model to fit the data, it is reassuring that
the residuals appear qualitatively uncorrelated in this figure.

We then sample the joint posterior probability (\eqalt{joint-post}) using
\emcee\ \citep{Goodman:2010, Foreman-Mackey:2013}.
We initialize \exampleinwalkers~walkers\footnote{\responseb{``Walker'' is the
name given by \citet{Goodman:2010} and \citet{Foreman-Mackey:2013} to each
parallel MCMC chain in the ensemble sampling algorithm used in \emcee.}} by
sampling from a three-dimensional isotropic Gaussian centered the MAP
parameter vector and with a standard deviation of $10^{-4}$ in each dimension.
We then run \exampleinburn~steps of burn-in and \exampleinsteps~steps of MCMC.
To assess convergence, we estimate the mean integrated autocorrelation time of
the chain across parameters \citep{Sokal:1989, Goodman:2010} and find that the
chain results in \exampleineff~effective samples.

@@

\subsection{Example 4: Asteroseismic oscillations}\sectlabel{astero}

@@

\newpage
\subsection{Code availability}

@@

For a GP kernel to be valid, it must produce a positive definite covariance
matrix for all input coordinates.
For stationary kernels, this is equivalent~--~by Bochner's theorem \citep[see
section 4.2.1 in][]{Rasmussen:2006}~--~to requiring that the kernel be the
Fourier transform of a positive finite measure.
This means that the power spectrum of a positive definite kernel must be
positive for all frequencies.
This result is intuitive because, since the power spectrum of a process is
defined as the expected squared amplitude of the Fourier transform of the time
series, it must be non-negative.
